The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-10-18 paper.qmd
changed format to jasa and edited reference so that it would work

{{< pagebreak >}}

# Introduction

@@

## Spatial example
There are far too many uncertainty visualisations to exhaustively discuss them all. 
Instead we focus on the changes made to a single plot, the choropleth map. 
Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
Utilising a single example will help isolate the ideas we are trying to convey. 
Therefore, it is important to remember that even though we focus our discussion on the choropleth map, the theoretical approach we outline in this review is useful to all uncertainty visualisations regardless of their application. 
Additionally, our examples focus on incorporating uncertainty through colour manipulation, as that is the key visual channel used in a choropleth map.
However, the methods we discuss go beyond variations in a colour palette.
Even though they are not explicitly shown, visualisations that depict uncertainty using layers such as position or shape, and more complicated graphics that incorporate animation or interactivity are also within the scope of this review.
We will use the choropleth maps as a tool to clearly highlight the costs and benefits of each approach.

@@

# Ignoring uncertainty
A good place to start might be at a deceptively straight forward question, why should we include uncertainty at all?

## The choropleth map
@fig-choropleth depicts a choropleth map of the counties of Iowa. 
Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend. 
The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while in the high variance, there is more variance within each county than between all the counties, so we should expect it to (at least in some capacity) overwhelm the spatial trend. 
Is this aspect of the data and the spatial data it communicates clear in in the map? 
Is the strength of the trend communicated through the visualisation?

@@

## Signal-suppression
Uncertainty visualisation is required for transparency. 
The two choropleth maps that appear to be identical in @fig-choropleth highlight the issues with simply electing to ignore uncertainty. 
This sentiment appears frequently in the uncertainty visualisation literature.
Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. 
Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. 
Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the validity of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. 
Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

@@

## Uncertainty as a signal
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss if we did not mention these alternative approaches. 
Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. 
Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as a variable of importance in of itself [@Blenkinsop2000]. 
While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as uncertainty at all. 
It is acting as signal.

@@

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
You may think "well, I will just add a dimension to my plot that includes uncertainty". 
This is a reasonable approach. 
The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
However, it is unclear if this approach is sufficient for our purposes.

## The bivariate map
@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional colour palette. 
In this graphic, temperature is still mapped to hue, but the variance is included by utilising colour saturation. 
While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. 
This means the uncertainty *is technically* being communicated, however the main message of the graphic is still the spatial trend (that may not exist). 
The graphic did not suppress the invalid signal, so it is not performing signal-suppression as we would like. 
At this point, it might be reasonable to ask, why? 
Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?

@@

## Why this approach may (or may not) work
The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. 
For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. 
Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006]. 
While this is part of the problem in uncertainty visualisation, it is not the complete picture. 
@fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppression. 
If we cannot treat uncertainty the same as we would any other variable, how should we treat it? 
We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.

### It's a variable... it's metadata... it's uncertainty?
Describing what uncertainty actually is, is surprisingly hard. 
Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty. 
Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. 
There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. 
Rather than trying to define uncertainty by looking at the myriad ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.

@@

### Visualising the "single integrated uncertain value"
Typically, when making visualisations, we want the visual channels to be separable. 
That is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. 
Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a single integrated channel. 
Visualising uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022]. 
We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.

@@

# Combining uncertainty and signal in a transformed space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves. 
As a matter of fact, some uncertainty visualisation authors already have.

## Value Suppressing Uncertainty Palettes
The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
Since the palette was designed with the extraction of individual values in mind and it has only been tested on simple value extraction tasks [@Correll2018] or search tasks [@Ndlovu2023], it is unclear how effective the method is at suppressing broader insights such as spatial trends.

@@

## What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. 
Specifically, uncertainty and the purpose of visualisation are somewhat at odds with one another. 
There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). 
Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals. 
If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. 
In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. 
However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. 
For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights. 
Unfortunately, it is unclear how reliably this would work.

### There is no uncertainty in EDA
Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty. 
Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing. 
In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis, which means there is no inference, and by extension, there is no uncertainty.

@@

### The limitations of explicitly visualising uncertainty and signal
The lack of versatility of the VSUP is easy to see with a simple example.

@@

# Implicitly Combining Uncertainty and Signal
Rather than trying to figure out how to combine signal and uncertainty into a single colour, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.

## Pixel map
@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each coloured using outcomes from the larger area's (i.e. the county's) average temperature sampling distribution. 
The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case. 
While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see. 
This means the graphic also achieves the third criteria for signal-suppression, i.e. our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.

@@

## Show me the data
To thoroughly understand why pixel maps seem to successfully perform signal suppression, we would need a more thorough discussion than what can be offered here. 
However, this does not prohibit us from discussing potential reasons for the graph's effectiveness. 
There is a second type of uncertainty that the pixel map quietly communicates but we have not touched on yet, and this uncertainty is also conveyed by visualisations of our raw data. 
That is, these graphics can show assumption violations.

@@

# Evaluating uncertainty visualisations
If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. 
Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", is lacking if we want any serious results.
However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].

@@

## Current methods
Including uncertainty in a visualisation comes with many secondary benefits.
Examples of these benefits include better decisions, more trust in the results and the ability to extract additional statistics, such as the variance.
Ultimately, these secondary benefits are not the primary goal of uncertainty visualisation, and evaluating uncertainty visualisations on these criteria often has unintended consequences.

### Value extraction of uncertainty statistics
Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and its variance [@Kinkeldey2014]. 
This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". 
This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? 
The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being uncertainty visualisations [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition. 
By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainty's ability to suppress statistics. 
This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.

@@

### Trust, confidence, and risk aversion
Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. 
Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018]. 
Science communication should be primarily concerned with accuracy. 
This accuracy may not always involve showing the data exactly as it is, for example we may want to adjust our graphic to correct for visual heuristics, but these changes should still be in pursuit of accuracy.
Setting trust as the variable of interest implicitly encourages statisticians to set trust as the primary goal of communication. 
Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. 
We can see this effect pop up in the results of evaluation studies. 
For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. 
Despite decreased trust being a desired outcome of signal suppression, the author's discussion implied this result was not desired. 
This perspective ends up extending to visualisation authors as well.
@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as untrustworthy. 
This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 
In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions. 
This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.
Of course, this does not mean measuring trust or perceived trustworthiness has no place in visualisation experiments.
After all, if participants political motivations can influence their ability to accurately read a plot [@nurse2020], then their level of trust in the information could have a similar effect.
However, it does mean trust should not be the primary goal of uncertainty visualisation evaluation experiments.

@@

### Questions that attempt to capture signal-suppression
There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. 
These studies try to measure the effect of some kind of single integrated value, but the methodology is often ad-hoc with varying levels of success.

@@

## Testing signal suppression
If we cannot ask direct questions about uncertainty, or measure the secondary effects of uncertainty, or ask indirect questions about the uncertainty; then how are we supposed to evaluate uncertainty visualisations? 
How do you measure something that disappears the second you look directly at it?
To evaluate uncertainty visualisations, we need an experimental design that evaluates uncertainty as noise, not as signal. 
This means we need to measure *uncertainty's impact on the signal*, not the uncertainty itself.

### Comparing to hypothesis tests
The most obvious way to evaluate uncertainty visualisations is to compare the visualisations to statistical tests. 
If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 
The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test. Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation. 
The lineup protocol is a good example of this approach.
Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
Lineups can also be used to evaluate the effectiveness of different types of plots [@Hofmann2012] and design decisions [@vanderplas2017].
The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 
@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. 
In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference. 
This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy. 
The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is the expected and we are looking for the unexpected.
This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.

@@

### Qualitative Studies
Alternatively visualisation research could shift away from the accuracy concept all together and ask questions that allow for open ended responses. This method can enlighten authors as to *how* the uncertainty information was used by the participants. 
@Hofmann2012 captured this by asking participants why they considered a particular plot to be more right-shifted, even though this qualitative analysis did not make it into the final paper. 
@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making. 
The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. 
They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc., and identified where in the decision making process participants made these considerations.

### Heuristics
While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression. 
Heuristic checks look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].

@@

# Conclusions and Future work
This paper has identified gaps in the uncertainty visualisation literature that must be filled for the field to progress. 
We formalised the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.

@@

# Acknowledgements
The first author of this paper is supported in part by a scholarship from the the Australian Energy Market Operator. 
The R packages were used for this work were: `tidyverse` [@tidyverse], `Vizumap` [@Vizumap], `RColorBrewer` [@RColorBrewer], `scales` [@scales], `sf` [@sf], `ggrepel` [@ggrepel], `urbnmapr` [@urbnmapr], `flextable` [@flextable], `colorspace` [@colorspace], and `rgeos` [@rgeos]. 
The GitHub repository for this paper can be found at https://github.com/harriet-mason/ARSA-UncertaintyLitReview which contains the files required to reproduce this article in full.

# Bibliography

>>>>

## Introduction

@@

### Spatial example
There are far too many uncertainty visualisations to exhaustively discuss them all. 
Instead we focus on the changes made to a single plot, the choropleth map. 
Due to the field's origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. 
Utilising a single example will help isolate the ideas we are trying to convey. 
Therefore, it is important to remember that even though we focus our discussion on the choropleth map, the theoretical approach we outline in this review is useful to all uncertainty visualisations regardless of their application. 
Additionally, our examples focus on incorporating uncertainty through colour manipulation, as that is the key visual channel used in a choropleth map.
However, the methods we discuss go beyond variations in a colour palette.
Even though they are not explicitly shown, visualisations that depict uncertainty using layers such as position or shape, and more complicated graphics that incorporate animation or interactivity are also within the scope of this review.
We will use the choropleth maps as a tool to clearly highlight the costs and benefits of each approach.

@@

## Ignoring uncertainty
A good place to start might be at a deceptively straight forward question, why should we include uncertainty at all?

### The choropleth map
@fig-choropleth depicts a choropleth map of the counties of Iowa. 
Each of these counties are coloured according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend. 
The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while in the high variance, there is more variance within each county than between all the counties, so we should expect it to (at least in some capacity) overwhelm the spatial trend. 
Is this aspect of the data and the spatial data it communicates clear in in the map? 
Is the strength of the trend communicated through the visualisation?

@@

### Signal-suppression
Uncertainty visualisation is required for transparency. 
The two choropleth maps that appear to be identical in @fig-choropleth highlight the issues with simply electing to ignore uncertainty. 
This sentiment appears frequently in the uncertainty visualisation literature.
Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. 
Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. 
Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the validity of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. 
Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

@@

### Uncertainty as a signal
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss if we did not mention these alternative approaches. 
Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. 
Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as a variable of importance in of itself [@Blenkinsop2000]. 
While uncertainty can provide useful information in decision making, it is important to recognise that the "uncertainty" in these cases is not acting as uncertainty at all. 
It is acting as signal.

@@

## Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. 
You may think "well, I will just add a dimension to my plot that includes uncertainty". 
This is a reasonable approach. 
The simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel. 
However, it is unclear if this approach is sufficient for our purposes.

### The bivariate map
@fig-bivariate depicts a variation of the choropleth map, where we have a two dimensional colour palette. 
In this graphic, temperature is still mapped to hue, but the variance is included by utilising colour saturation. 
While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. 
This means the uncertainty *is technically* being communicated, however the main message of the graphic is still the spatial trend (that may not exist). 
The graphic did not suppress the invalid signal, so it is not performing signal-suppression as we would like. 
At this point, it might be reasonable to ask, why? 
Why is including the uncertainty as a variable insufficient to achieve signal-suppression, and what changes should we make to ensure signal-suppression occurs?

@@

### Why this approach may (or may not) work
The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. 
For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. 
Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006]. 
While this is part of the problem in uncertainty visualisation, it is not the complete picture. 
@fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppression. 
If we cannot treat uncertainty the same as we would any other variable, how should we treat it? 
We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.

#### It's a variable... it's metadata... it's uncertainty?
Describing what uncertainty actually is, is surprisingly hard. 
Most authors simply avoid the problem and describe the characteristics of uncertainty, of which there are plenty. 
Often, uncertainty is split using an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. 
There are enough qualitative descriptors of uncertainty to fill a paper, but, none of this is particularly helpful in understanding how to integrate it into a visualisation. 
Rather than trying to define uncertainty by looking at the myriad ways in which it *does* appear in an analysis, we may find it easier to look at where it *does not*.

@@

#### Visualising the "single integrated uncertain value"
Typically, when making visualisations, we want the visual channels to be separable. 
That is, we don't want the data represented through one visual channel to interfere with the others [@Smart2019]. 
Mapping uncertainty and signal to separable channels allows them to be read separately, which does not align with the goal of communicating them as a single integrated channel. 
Visualising uncertainty and signal separately allows the uncertainty information to simply be ignored, which is a pervasive issue in current uncertainty visualisation methods [@uncertchap2022]. 
We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" as we read the signal and the uncertainty separately.

@@

## Combining uncertainty and signal in a transformed space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves. 
As a matter of fact, some uncertainty visualisation authors already have.

### Value Suppressing Uncertainty Palettes
The Value Suppressing Uncertainty Palette (VSUP) [@Correll2018], was designed with the intention of preventing high uncertainty values from being extracted from a map.
Since the palette was designed with the extraction of individual values in mind and it has only been tested on simple value extraction tasks [@Correll2018] or search tasks [@Ndlovu2023], it is unclear how effective the method is at suppressing broader insights such as spatial trends.

@@

### What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. 
Specifically, uncertainty and the purpose of visualisation are somewhat at odds with one another. 
There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). 
Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation and using it to extract several signals. 
If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. 
In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. 
However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. 
For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights. 
Unfortunately, it is unclear how reliably this would work.

#### There is no uncertainty in EDA
Earlier we established that uncertainty is a by-product of inference, which means without inference, there is no uncertainty. 
Often EDA is used to give us an understanding of our data and identify which signals are worth pursuing. 
In this sense, EDA is the visual parallel to descriptive statistics, as it is performed without an explicit hypothesis, which means there is no inference, and by extension, there is no uncertainty.

@@

#### The limitations of explicitly visualising uncertainty and signal
The lack of versatility of the VSUP is easy to see with a simple example.

@@

## Implicitly Combining Uncertainty and Signal
Rather than trying to figure out how to combine signal and uncertainty into a single colour, we can just display a sample instead and allow the viewer to extract *both* the estimate and the variance.

### Pixel map
@fig-pixel displays a pixel map [@Lucchesi2021], which is a variation of the choropleth map where each area is divided up into several smaller areas, each coloured using outcomes from the larger area's (i.e. the county's) average temperature sampling distribution. 
The spatial trend is clearly visible in the low variance case, but functionally disappears in the low variance case. 
While the spatial trend is just barely visible in the high uncertainty case, it is much harder to see. 
This means the graphic also achieves the third criteria for signal-suppression, i.e. our difficulty in seeing the distribution is proportional to the level of uncertainty in the graphic.

@@

### Show me the data
To thoroughly understand why pixel maps seem to successfully perform signal suppression, we would need a more thorough discussion than what can be offered here. 
However, this does not prohibit us from discussing potential reasons for the graph's effectiveness. 
There is a second type of uncertainty that the pixel map quietly communicates but we have not touched on yet, and this uncertainty is also conveyed by visualisations of our raw data. 
That is, these graphics can show assumption violations.

@@

## Evaluating uncertainty visualisations
If we want to make conclusions about how effective any uncertainty visualisation method is we need to look at the results of evaluation experiments. 
Unfortunately the illustrative methods we have used thus far, i.e. showing a graphic and saying "wow look at this", is lacking if we want any serious results.
However, despite the abundance of uncertainty visualisation evaluation experiments, existing literature reviews have struggled to synthesise them into any common rules [@Kinkeldey2014; @Hullman2016].

@@

### Current methods
Including uncertainty in a visualisation comes with many secondary benefits.
Examples of these benefits include better decisions, more trust in the results and the ability to extract additional statistics, such as the variance.
Ultimately, these secondary benefits are not the primary goal of uncertainty visualisation, and evaluating uncertainty visualisations on these criteria often has unintended consequences.

#### Value extraction of uncertainty statistics
Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and its variance [@Kinkeldey2014]. 
This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". 
This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? 
The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being uncertainty visualisations [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition. 
By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainty's ability to suppress statistics. 
This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.

@@

#### Trust, confidence, and risk aversion
Trust is a by-product of displaying uncertainty so it is commonly measured in uncertainty evaluation studies [@Hullman2019]. 
Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; @ONeill2018]. 
Science communication should be primarily concerned with accuracy. 
This accuracy may not always involve showing the data exactly as it is, for example we may want to adjust our graphic to correct for visual heuristics, but these changes should still be in pursuit of accuracy.
Setting trust as the variable of interest implicitly encourages statisticians to set trust as the primary goal of communication. 
Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. 
We can see this effect pop up in the results of evaluation studies. 
For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. 
Despite decreased trust being a desired outcome of signal suppression, the author's discussion implied this result was not desired. 
This perspective ends up extending to visualisation authors as well.
@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as untrustworthy. 
This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 
In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions. 
This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.
Of course, this does not mean measuring trust or perceived trustworthiness has no place in visualisation experiments.
After all, if participants political motivations can influence their ability to accurately read a plot [@nurse2020], then their level of trust in the information could have a similar effect.
However, it does mean trust should not be the primary goal of uncertainty visualisation evaluation experiments.

@@

#### Questions that attempt to capture signal-suppression
There is a collection of studies that seem to be aware of the issues behind using trust or value extraction to evaluate uncertainty visualisations. 
These studies try to measure the effect of some kind of single integrated value, but the methodology is often ad-hoc with varying levels of success.

@@

### Testing signal suppression
If we cannot ask direct questions about uncertainty, or measure the secondary effects of uncertainty, or ask indirect questions about the uncertainty; then how are we supposed to evaluate uncertainty visualisations? 
How do you measure something that disappears the second you look directly at it?
To evaluate uncertainty visualisations, we need an experimental design that evaluates uncertainty as noise, not as signal. 
This means we need to measure *uncertainty's impact on the signal*, not the uncertainty itself.

#### Comparing to hypothesis tests
The most obvious way to evaluate uncertainty visualisations is to compare the visualisations to statistical tests. 
If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. 
The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test. Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation. 
The lineup protocol is a good example of this approach.
Lineups are a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. 
Lineups can also be used to evaluate the effectiveness of different types of plots [@Hofmann2012] and design decisions [@vanderplas2017].
The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. 
@Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. 
In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference. 
This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy. 
The connection with visualisation and statistical tests even extends to EDA, where @Hullman2021 commented that a visualisation can be considered a model check where the null hypothesis is the expected and we are looking for the unexpected.
This is all to say that comparing visualisations to hypothesis tests is not uncharted territory.

@@

#### Qualitative Studies
Alternatively visualisation research could shift away from the accuracy concept all together and ask questions that allow for open ended responses. This method can enlighten authors as to *how* the uncertainty information was used by the participants. 
@Hofmann2012 captured this by asking participants why they considered a particular plot to be more right-shifted, even though this qualitative analysis did not make it into the final paper. 
@Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making. 
The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. 
They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc., and identified where in the decision making process participants made these considerations.

#### Heuristics
While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression. 
Heuristic checks look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].

@@

## Conclusions and Future work
This paper has identified gaps in the uncertainty visualisation literature that must be filled for the field to progress. 
We formalised the uncertainty visualisation problem, and in doing so highlighted an untouched area of uncertainty visualisation research.

@@

## Acknowledgements

The first author of this paper is supported in part by a scholarship from the the Australian Energy Market Operator. 
The R packages were used for this work were: `tidyverse` [@tidyverse], `Vizumap` [@Vizumap], `RColorBrewer` [@RColorBrewer], `scales` [@scales], `sf` [@sf], `ggrepel` [@ggrepel], `urbnmapr` [@urbnmapr], `flextable` [@flextable], `colorspace` [@colorspace], and `rgeos` [@rgeos]. 
The GitHub repository for this paper can be found at https://github.com/harriet-mason/ARSA-UncertaintyLitReview which contains the files required to reproduce this article in full.

## Bibliography
